Modelling $q_{95}$ Windows for the Suppression of Edge Localized Modes by Resonant Magnetic Perturbations in the DIII-D Tokamak

### 2020-07-02 ExperimentalData/DIIID/145380/Paper/paper.tex
Updates to paper

Numerical simulations made using the cylindrical,  nonlinear, two-fluid, reduced-magneto\-hydro\-dynamical (MHD), initial-value code, TM1\,\cite{tm1,tm2,tm3} have shed considerable light on the hitherto poorly understood  physical mechanism that underlies RMP-induced ELM
suppression in H-mode tokamak discharges.\cite{hu} The simulations  in question make a plausible case  that the density  pump-out  phenomenon is associated with the formation of   locked (i.e., non-rotating) helical magnetic
island chains at the bottom of the pedestal, whereas the ELM suppression phenomenon is associated with the formation of a locked helical magnetic island chain at the
top of the pedestal. The prevailing hypothesis is that such an island chain suppresses ELMs by limiting the expansion of the
pedestal, and, thereby, preventing it from attaining a width sufficient to destabilize peeling-ballooning modes\,\cite{connor}  (which are thought to trigger ELMs).\cite{d3d}

Recently, a toroidal generalization of the 
cylindrical asymptotic matching model presented
in Ref.~\onlinecite{rf1} was formulated and
used to model  RMP-induced ELM suppression experiments performed on the DIII-D tokamak, \cite{rf2}
leading to similar conclusions to the aforementioned TM1
studies. The primary aim of this paper is to employ this
new model to try to account for the  $q_{95}$ ELM
suppression windows  that are apparent when the edge safety-factor is slowly ramped in
a particular DIII-D discharge in which an $n=3$
RMP is used to control ELMs.\cite{d3d,d3d2}

It is well-known that magnetic reconnection caused when a  (stable) tearing mode is driven by a static RMP that is resonant at a particular magnetic flux-surface in a tokamak plasma is
facilitated when the associated ``natural frequency'' is relatively small.\cite{rfa,rfb}  The natural frequency  of a  (stable) tearing mode is the
helical phase velocity that the mode would possess were it naturally unstable (in the absence of the RMP).  Driven magnetic reconnection leads to the
formation of a locked magnetic island chain at the resonant surface in question. 
Hence, the prevailing hypothesis is that a $q_{95}$ window for
RMP-induced ELM suppression occurs when $q_{95}$ is such that the natural frequency of a tearing mode resonant  at 
the top of the pedestal is close to zero. 
Unfortunately, there is currently some uncertainty in the fusion community regarding the form of the natural frequency.

According to
linear tearing mode theory, a tearing mode is essentially convected by
the local electron fluid at the resonant surface (see the discussion at the end of the Appendix).\cite{hender,cole} Hence, if the response of an H-mode tokamak plasma to an applied RMP
is governed by linear physics then we would expect the
natural frequency to take the form\,\cite{lin1,lin2,lin3}
\begin{equation}
\varpi_{\rm linear} = - n\,(\omega_E+\omega_{\ast\,e}),\label{e1}
\end{equation}
where
\begin{align}
\omega_E &= -\frac{d{\mit\Phi}}{d{\mit\Psi}_p},\\[0.5ex]
\omega_{\ast\,e} &=\frac{T_e}{e}\,\frac{d\ln p_e}{d{\mit\Psi}_p}.
\end{align}
Here, ${\mit\Psi}_p$ is the equilibrium poloidal magnetic
flux (divided by $2\pi$), ${\mit\Phi}({\mit\Psi}_p)$ the equilibrium scalar
electric potential, $p_e({\mit\Psi}_p)$ the equilibrium
electron pressure, $T_e({\mit\Psi}_p)$ the equilibrium
electron temperature, $e$ the magnitude of the electron
charge, and $n$ the toroidal mode number of the RMP. 
Moreover, the right-hand side of Eq.~(\ref{e1}) is evaluated
at the ``rational''  (i.e., resonant) magnetic flux-surface at which the
safety-factor
\begin{equation}
q({\mit\Psi}_p)= \frac{d{\mit\Psi}_t}{d{\mit\Psi}_p},
\end{equation}
takes the rational value $n/m$, where $m$ is a positive integer. Here, $m$ and $n$ are the numbers of poloidal and
toroidal periods, respectively, of the helical magnetic island chain driven at the rational surface.  Furthermore, ${\mit\Psi}_t({\mit\Psi}_p)$ is the equilibrium toroidal magnetic flux (divided by $2\pi$).

According to nonlinear tearing mode theory, a tearing mode is essentially convected by
the local ion fluid at the rational surface (again, see the discussion at the end of the Appendix).\cite{nl1,nl2,nl3}
In fact, if the response of an H-mode tokamak plasma to an applied RMP
is governed by nonlinear physics then we would expect the
natural frequency to take the form\,\cite{rf2}
\begin{equation}\label{e2}
\varpi_{\rm nonlinear} = -n\left(\omega_E +\left[1-L_{00}^{\,ii}+L_{01}^{\,ii}\left(\frac{\eta_i}{1+\eta_i}\right)\right]
\omega_{\ast\,i}
%\right.\nonumber\\[0.5ex]&\left.
-\left[L_{00}^{\,iI}- L_{01}^{\,iI}\left(\frac{\eta_I}{1+\eta_I}\right)\right]\omega_{\ast\,I}\right),
\end{equation}
where
\begin{align}
\omega_{\ast\,a}({\mit\Psi}_p) &=-\frac{T_a}{Z_a\,e}\,\frac{d\ln p_a}{d{\mit\Psi}_p},\\[0.5ex]
\eta_a({\mit\Psi}_p)&=\frac{d\ln T_a}{d\ln n_a},
\end{align}
for $a=i, I$. 
Here,  $Z_i$, $n_i({\mit\Psi}_p)$, $T_i({\mit\Psi}_p)$, and $p_i({\mit\Psi}_p)=n_i\,T_i$   are the charge number, equilibrium number density, 
equilibrium temperature, and equilibrium pressure
of the majority (thermal) ions, respectively,
whereas  $Z_I$, $n_I$, $T_I$, $p_I = n_I\,T_I$
are the corresponding quantities for the impurity ions.  
Furthermore, $L_{00}^{ii}({\mit\Psi}_p)$, $L_{01}^{ii}({\mit\Psi}_p)$,  $L_{00}^{II}({\mit\Psi}_p)$, and $L_{01}^{II}({\mit\Psi}_p)$ are neoclassical parameters that are defined
in Ref.~\onlinecite{rf2}. Note that these parameters are
affected by charge exchange  with neutrals. Finally,
as before, the right-hand side of Eq.~(\ref{e2}) is evaluated
at the rational magnetic flux-surface.

@@

The secondary aim of this paper is to determine which of the three aforementioned choices for the natural frequency can best account for the   $q_{95}$ ELM
suppression windows  observed in DIII-D.

\section{Description of Theoretical Model}
The theoretical model of the response of a tokamak plasma to an externally applied RMP that is employed in this
paper is described in detail in Ref.~\onlinecite{rf2}. The model employs a standard asymptotic matching approch.\cite{fkr,am1,am2,am3}
According to this approach, the response of the plasma to the applied RMP is governed by a combination of flux-freezing and
perturbed force balance (this combination is often referred to as ``marginally-stable ideal-MHD", which is a misnomer because MHD {\em per se}\/ plays no
role) everywhere in the plasma apart from a number of relatively narrow (in the radial direction) regions in which the applied
perturbation resonates with the equilibrium magnetic field. Magnetic reconnection can take place within the resonant regions to
produce relatively thin magnetic islands. Within the resonant regions, the plasma response is governed by nonlinear,
as opposed to linear, two-fluid resistive MHD. This is the case because the widths of the magnetic island chains
driven at the resonant surfaces exceed the linear layer widths (which invalidates linear theory).\cite{rf1}
Thus, when employing the asymptotic matching approach, the equations of flux-freezing and perturbed force balance
are solved in the so-called ``outer region'' that comprises most of the plasma (and the surrounding vacuum), the equations of 
nonlinear two-fluid resistive-MHD are solved in the various resonant layers that constitute the so-called ``inner region'', and the two
sets of solutions are then asymptotically matched to one another.

A toroidal tokamak equilibrium exhibits two distinct types of response to an applied RMP.\cite{paz1,kink0,kink1}
The first of these is known as the ``tearing response''---this is a non-ideal-MHD response that is associated with the
formation of current sheets and magnetic island chains at various resonant surfaces within the plasma. The second
response type is known as the ``kink response''---this is an edge-localized ideal-MHD response that is associated
with coupling to a stable non-resonant kink mode. For the case of the tearing response, our model employs an
approximation in which the plasma response is assumed to be vacuum-like between the various resonant surfaces. On the other hand,
the kink response of the plasma is calculated using the {\tt GPEC} code.\cite{ipec,gpec}

Our model has been implemented in the {\tt EPEC} (Extended Perturbed Equilibrium Code) code. The name
of this code reflects the fact that the nonlinear evolution of tearing modes in a
tokamak plasma has far more in common with the $1\tfrac{1}{2}$-D
evolution of the global plasma equilibrium than it does with
conventional linear tearing mode physics. (See the discussion in the Appendix.) In particular, the Alfv\'{e}n time is
an irrelevant timescale in nonlinear tearing mode theory, and
is also  very much shorter than the timescales on which
physical quantities of interest actually evolve. (Note that all timescales are normalized to the Alfv\'{e}n time
in Ref.~\onlinecite{rf2}. However, this is just a matter of convention. With the benefit of hindsight, it would
have been better to normalize the timescales with respect to a diamagnetic timescale, in which case the Alfv\'{e}n time would have completely dropped
out of the final system of equations.)

\section{Modeling of DIII-D Discharge \#145380}

@@

\bibitem{nl1} R. Fitzpatrick, and F.L. Waelbroeck, Phys.\ Plasmas {\bf 12}, 022307 (2005).

\bibitem{nl2} R. Fitzpatrick, Phys.\ Plasmas {\bf 25}, 042503 (2018).

\bibitem{nl3} R. Fitzpatrick, Phys.\ Plasmas {\bf 25}, 112505 (2018).

@@

@@

@@

\appendix
\section{Linear Versus Nonlinear Tearing Mode Theory in Tokamak Plasmas}\label{appa}
The discussion in this Appendix outlines certain facts pertaining
to tearing modes in tokamak plasmas that ought to be common knowledge in the 
magnetic fusion community (but appear not to be).

Consider a conventional tokamak plasma equilibrium. To lowest order, the equilibrium is governed by force balance,
\begin{equation}\label{ea}
-\nabla p + {\bf j}\times {\bf B} \simeq {\bf 0},
\end{equation}
and
flux freezing,
\begin{equation}\label{eb}
{\bf E} + {\bf V}\times {\bf B} \simeq {\bf 0}.
\end{equation}
This is the case because the other terms in the plasma equation of motion,
(\ref{ea}), and the plasma Ohm's law, (\ref{eb}), are much
smaller in magnitude than the retained terms. (Here, $p$ is the
scalar pressure, ${\bf j}$ the current density, ${\bf B}$ the
magnetic field strength, ${\bf E}$ the electric field strength, and ${\bf V}$ the plasma velocity). If we average the full plasma equation of motion and the full Ohm's law, as well as the full continuity and energy evolution equations, over magnetic flux-surfaces
then the dominant terms in these equations are annihilated, and
we end up with a set of $1\tfrac{1}{2}$-D evolution equations,
according to which the magnetic field and plasma current
evolve on the relatively long resistive timescale, $\tau_R$,
whereas the density and temperature evolve on a somewhat
shorter transport timescale.\cite{transp}

A linear tearing mode is a helical instability of a tokamak plasma equilibrium that is related to a shear-Alfv\'{e}n wave.\cite{haz} The mode resonates with the plasma at
the so-called rational magnetic flux-surface at which the
shear-Alfv\'{e}n velocity is zero. If $q$ is the safety-factor, $m$ the poloidal mode number, and $n$ the
toroidal mode number, then the rational
flux-surface is characterized by $q=m/n$. A thin resistive layer forms
around the rational surface that permits the reconnection of magnetic flux on a much faster timescale than the global
resistive evolution timescale, $\tau_R$, resulting in the formation of a helical magnetic island chain. Let the Alfv\'{e}n time,
$\tau_A$, be the
typical timescale on which a (compressional) Alfv\'{e}n
wave traverses the plasma.
In conventional tokamak plasmas, $\tau_A\ll \tau_R$. In fact,
the magnetic Lundquist number, $S\equiv \tau_R/\tau_A$,  typically exceeds $10^{\,8}$ in present-day tokamak plasmas,
and will likely exceed $10^{\,10}$ in ITER plasmas.
 It turns out that the width of the resistive layer is typically a factor $S^{\,-2/5}$ smaller than the
plasma minor radius, whereas magnetic flux is reconnected on
the hybrid timescale $\tau_A\ll \tau_A^{2/5}\,\tau_R^{3/5}\ll \tau_R$.\cite{fkr} Note that  the
structure of the tearing mode  outside the resistive layer is simply governed by a combination of flux-freezing and perturbed force balance.

@@

Figure~\ref{fig14} also sketches the
typical plasma current and vorticity patterns in the nonlinear regime. Observe that there is now a considerable distinction between an X-point and an O-point. The reason for this is that the island is sufficiently wide that the plasma cannot diffuse across it resistively sufficiently rapidly to
avoid being trapped inside the magnetic separatrix. Indeed, both the regions inside and outside the magnetic separatrix
are governed by a combination of flux-freezing and perturbed force balance, whereas the resistive
layer, within which the plasma can easily slip through the magnetic field,  is diverted onto the magnetic separatrix. Moreover, as
the island grows, the width of the separatrix layer becomes
increasing negligible with respect to the island width. It
turns out that the separatrix layer does not significantly affect the evolution of a nonlinear magnetic island chain. Indeed,
the celebrated Rutherford island width evolution equation,
which governs the evolution of the island width, is simply a
$1\tfrac{1}{2}$-D evolution equation for a helical magnetic
equilibrium localized around the rational surface.\cite{ruth}
The Alfv\'{e}n time plays no role in this evolution (any more
than it plays a role in the $1\tfrac{1}{2}$-D evolution of
the global magnetic field). Likewise, the equation that
determines the phase velocity of a nonlinear magnetic island chain (see Ref.~\onlinecite{nl2}, and references therein) is essentially an
expression of the $1\tfrac{1}{2}$-D evolution of the plasma
flow profiles across the island region. As before, the  Alfv\'{e}n time plays no role in this evolution (any more
than it plays a role in the $1\tfrac{1}{2}$-D evolution of
the global plasma flow profiles).

>>>>

Numerical simulations made using the cylindrical,  nonlinear, two-fluid, reduced-magneto\-hydro\-dynamical (MHD), initial-value code, TM1\,\cite{tm1,tm2,tm3} have shed considerable light on the hitherto poorly understood  physical mechanism that underlies RMP-induced ELM
suppression in H-mode tokamak discharges.\cite{hu} The simulations  in question make a plausible case  that the density  pump-out  phenomenon is associated with the formation of   locked (i.e., non-rotating) helical magnetic
island chains at the bottom of the pedestal, whereas the ELM-suppression phenomenon is associated with the formation of a locked helical magnetic island chain at the
top of the pedestal. The prevailing hypothesis is that such an island chain suppresses ELMs by limiting the expansion of the
pedestal, and, thereby, preventing it from attaining a width sufficient to destabilize peeling-ballooning modes\,\cite{connor}  (which are thought to trigger ELMs).\cite{d3d}

Recently, a toroidal generalization of the 
cylindrical asymptotic matching model presented
in Ref.~\onlinecite{rf1} was formulated and
used to model  RMP-induced ELM-suppression experiments performed on the DIII-D tokamak, \cite{rf2}
leading to similar conclusions to the aforementioned TM1
studies. The primary aim of this paper is to employ this
new model to try to account for the  $q_{95}$
ELM-suppression windows  that are apparent when the edge safety-factor is slowly ramped in
a particular DIII-D discharge in which an $n=3$
RMP is used to control ELMs.\cite{d3d,d3d2}

It is well-known that magnetic reconnection driven when a  (stable) tearing mode interacts with a static RMP that is resonant at a particular magnetic flux-surface in a tokamak plasma is
facilitated when the associated ``natural frequency'' is relatively small.\cite{rfa,rfb}  The natural frequency  of a  (stable) tearing mode is the
helical phase velocity that the mode would possess were it naturally unstable (in the absence of the RMP).  Driven magnetic reconnection leads to the
formation of a locked magnetic island chain at the resonant surface in question. 
Hence, the prevailing hypothesis is that a $q_{95}$ window for
RMP-induced ELM suppression occurs when $q_{95}$ is such that the natural frequency of a tearing mode resonant  at 
the top of the pedestal is close to zero. 
However, there is currently some uncertainty in the fusion community regarding the form of the natural frequency.

According to
linear tearing mode theory, a tearing mode is essentially convected by
the local electron fluid at the resonant surface (see the discussion in Sect.~\ref{natural}).\cite{hender,cole} Hence, if the response of an H-mode tokamak plasma to an applied RMP
is governed by linear physics then we would expect the
natural frequency to take the form\,\cite{lin1,lin2,lin3}
\begin{equation}
\varpi_{\rm linear} = - n\,(\omega_E+\omega_{\ast\,e}),\label{e1}
\end{equation}
where
\begin{align}
\omega_E &= -\frac{d{\mit\Phi}}{d{\mit\Psi}_p},\\[0.5ex]
\omega_{\ast\,e} &=\frac{T_e}{e}\,\frac{d\ln p_e}{d{\mit\Psi}_p}.
\end{align}
Here, ${\mit\Psi}_p$ is the equilibrium poloidal magnetic
flux (divided by $2\pi$), ${\mit\Phi}({\mit\Psi}_p)$ the equilibrium scalar
electric potential, $p_e({\mit\Psi}_p)$ the equilibrium
electron pressure, $T_e({\mit\Psi}_p)$ the equilibrium
electron temperature, $e$ the magnitude of the electron
charge, and $n$ the toroidal mode number of the RMP. 
Moreover, the right-hand side of Eq.~(\ref{e1}) is evaluated
at the ``rational''  (i.e., resonant) magnetic flux-surface at which the
safety-factor
\begin{equation}
q({\mit\Psi}_p)= \frac{d{\mit\Psi}_t}{d{\mit\Psi}_p},
\end{equation}
takes the rational value $n/m$, where $m$ is a positive integer. Here, $m$ and $n$ are the numbers of poloidal and
toroidal periods, respectively, of the helical magnetic island chain driven at the rational surface.  Furthermore, ${\mit\Psi}_t({\mit\Psi}_p)$ is the equilibrium toroidal magnetic flux (divided by $2\pi$).

According to nonlinear tearing mode theory, a tearing mode is essentially convected by
the local ion fluid at the rational surface (again, see the discussion in Sect.~\ref{natural}).\cite{nl1,nl2,nl3}
In fact, if the response of an H-mode tokamak plasma to an applied RMP
is governed by nonlinear physics then we would expect the
natural frequency to take the form\,\cite{rf2}
\begin{equation}\label{e2}
\varpi_{\rm nonlinear} = -n\left(\omega_E +\left[1-L_{00}^{\,ii}+L_{01}^{\,ii}\left(\frac{\eta_i}{1+\eta_i}\right)\right]
\omega_{\ast\,i}
%\right.\nonumber\\[0.5ex]&\left.
-\left[L_{00}^{\,iI}- L_{01}^{\,iI}\left(\frac{\eta_I}{1+\eta_I}\right)\right]\omega_{\ast\,I}\right),
\end{equation}
where
\begin{align}
\omega_{\ast\,a}({\mit\Psi}_p) &=-\frac{T_a}{Z_a\,e}\,\frac{d\ln p_a}{d{\mit\Psi}_p},\\[0.5ex]
\eta_a({\mit\Psi}_p)&=\frac{d\ln T_a}{d\ln n_a},
\end{align}
for $a=i, I$. 
Here,  $Z_i$, $n_i({\mit\Psi}_p)$, $T_i({\mit\Psi}_p)$, and $p_i({\mit\Psi}_p)=n_i\,T_i$   are the charge number, equilibrium number density, 
equilibrium temperature, and equilibrium pressure
of the majority (thermal) ions, respectively,
whereas  $Z_I$, $n_I$, $T_I$, $p_I = n_I\,T_I$
are the corresponding quantities for the impurity ions.  
Furthermore, $L_{00}^{ii}({\mit\Psi}_p)$, $L_{01}^{ii}({\mit\Psi}_p)$,  $L_{00}^{II}({\mit\Psi}_p)$, and $L_{01}^{II}({\mit\Psi}_p)$ are neoclassical parameters that are defined
in Ref.~\onlinecite{rf2}. Note that these parameters are
affected by charge exchange  with neutrals. Finally,
as before, the right-hand side of Eq.~(\ref{e2}) is evaluated
at the rational magnetic flux-surface.

@@

The secondary aim of this paper is to determine which of the three aforementioned choices for the natural frequency can best account for the   $q_{95}$ 
ELM-suppression windows  observed in DIII-D discharge \#145380.

\section{Description of Theoretical Model}
The theoretical model of the response of a tokamak plasma to an externally applied RMP that is used in this
paper is described in detail in Ref.~\onlinecite{rf2}. The model employs a standard asymptotic matching approach.\cite{fkr,am1,am2,am3}
According to this approach, the response of the plasma to the applied RMP is governed by a combination of flux-freezing and
perturbed force balance (this combination is often referred to as ``marginally-stable ideal-MHD", which is a misnomer because MHD {\em per se}\/ plays no
role) everywhere in the plasma apart from a number of relatively narrow (in the radial direction) regions in which the applied
perturbation resonates with the equilibrium magnetic field. Magnetic reconnection can take place within the resonant regions to
produce relatively thin magnetic islands. Within the resonant regions, the plasma response is governed by nonlinear,
as opposed to linear, two-fluid resistive MHD. This is the case because the widths of the magnetic island chains
driven at the resonant surfaces exceed the linear layer widths (which invalidates linear theory).\cite{rf1}
Thus, when employing the asymptotic matching approach, the equations of flux-freezing and perturbed force balance
are solved in the so-called ``outer region'' that comprises most of the plasma (and the surrounding vacuum), the equations of 
nonlinear two-fluid resistive-MHD are solved in the various resonant layers that constitute the so-called ``inner region'', and the two
sets of solutions are then asymptotically matched to one another.

A toroidal tokamak equilibrium exhibits two distinct types of response to an applied RMP.\cite{paz1,kink0,kink1}
The first of these is known as the ``tearing response''---this is a non-ideal-MHD response that is associated with the
formation of current sheets and magnetic island chains at various resonant surfaces within the plasma. The second
response type is known as the ``kink response''---this is an edge-localized ideal-MHD response that is associated
with coupling to a stable non-resonant kink mode. For the case of the tearing response, our model employs an
approximation in which the plasma response is assumed to be vacuum-like between the various resonant surfaces. On the other hand,
the kink response of the plasma is calculated exactly using the {\tt GPEC} code.\cite{ipec,gpec}

Our model has been implemented in the {\tt EPEC} (Extended Perturbed Equilibrium Code) code. The name
of this code reflects the fact that the nonlinear evolution of tearing modes in a
tokamak plasma has far more in common with the $1\tfrac{1}{2}$-D
evolution of the global plasma equilibrium than it does with
conventional linear tearing mode physics. (See the discussion in Sect.~\ref{general}.) In particular, the Alfv\'{e}n time is
an irrelevant timescale in nonlinear tearing mode theory, and
is also  very much shorter than the timescales on which
physical quantities of interest actually evolve. (Note that all timescales are normalized to the Alfv\'{e}n time
in Ref.~\onlinecite{rf2}. However, this is just a matter of convention. With the benefit of hindsight, it would
have been better to normalize the timescales with respect to a diamagnetic timescale, in which case the Alfv\'{e}n time would have completely dropped
out of the final system of equations.)

\section{Modeling of DIII-D Discharge \#145380}
\subsection{Plasma Equilibrium}
DIII-D discharge \#145380 is an ITER Similar Shape (ISS), ELMing, H-mode discharge with a toroidal magnetic field $B_T=-1.9\,{\rm T}$ in which the
plasma current $I_p$ is slowly ramped over a 2 s interval to scan the magnetic safety-factor.\cite{d3d,d3d2} The majority ions are Deuterium, whereas the
minority ions are Carbon~VI.
Figure~\ref{fig1} gives an overview of this discharge. Note that the pedestal density is relatively high ($n_{e\,{\rm ped}}\sim 3\times 10^{19}\,{\rm m}^{-3}$). 
A static RMP is applied to the plasma by running steady $n=3$ currents through the I-coil system.\cite{icoil}
Three windows of ELM suppression are evident: the first extends from $t=2840$--$2980$ ms; the
second extends from $t=3320$--$3560$ ms; and the third extends from $t=3880$--$4200$ ms.

Figure~\ref{fig2} shows the measured plasma equilibrium at the start of the current ramp ($t=2500$ ms). Likewise,
Fig.~\ref{fig3} shows the measured plasma profiles at the start of the current ramp. Here, ${\mit\Psi}_N$ is a
normalized poloidal magnetic flux defined such that the magnetic axis corresponds to ${\mit\Psi}_N=0$, 
and the last closed magnetic flux-surface (LCFS) to ${\mit\Psi}_N=1$. The perpendicular
diffusivity data comes from {\tt TRANSP} modelling.\cite{transp}

Measured plasma equilibrium and
plasma profile data is available every $50$ ms. The equilibrium and profile data at a given point in time is interpolated from
the measured data. In fact, quantities that depend on the plasma equilibrium and profiles are recalculated every 1 ms.  Data from the
{\tt GPEC} code is available every 100 ms. As before, the {\tt GPEC} data at a given point in time is interpolated from the
available {\tt GPEC} data, and any quantity that depends on this data is recalculated every 1 ms. There is one major exception to
the aforementioned scheme. The theoretical model used in this paper actually requires plasma equilibria and profiles that are unaffected
by the applied RMP. However, there is clear evidence that the equilibria and the profiles (in particular) are significantly modified
within the three ELM-suppression windows (presumably because wide magnetic island chains are driven in the pedestal). 
Hence, we do not use the equilibrium, profile, or {\tt GPEC}, data within the three ELM-suppression windows. Instead, we
interpolate across these windows using data taken on either side of the windows.

Figure~\ref{fig4} shows the safety-factor in DIII-D discharge \#145380 as a function of time. It can be seen
that $q_{95}\equiv q({\mit\Psi}_N=0.95)$ varies from about 4.1 at the start of the current ramp to about 3.3
at the end. However, the safety-factor at the magnetic axis, $q_0 \equiv q({\mit\Psi}_N=0.0)$, remains approximately
constant at about 1.0. For the sake of practicality,  the {\tt GPEC} and {\tt EPEC} calculations described in this paper ignore rational
surfaces that lie very close to the LCFS: in other words,  surfaces that lie in the region $0.997 < {\mit\Psi}_N\leq 1.000$.

Note that every quantity used in our theoretical model is derived from experimental measurements taken in DIII-D discharge \#145380, with one
exception. The exception is the neutral particle data. Unfortunately, no neutral particle data is available for DIII-D discharge \#145380.
Hence, we employ neutral particle data derived from a much earlier discharge (see Ref.~\onlinecite{rf2}).\cite{neutral}
In future, data from the new {\tt LLAMA} (Lyman Alpha Measurement Apparatus) system will be available in DIII-D discharges.\cite{llama}
Combining this data  with modeling from the SOLPS-ITER code\,\cite{solps} ought to generate sufficient information to fully determine the
neutral particle parameters required by the {\tt EPEC} code.

\subsection{Case 1: $\varpi=\varpi_{\rm nonlinear}$}
This first case that we shall consider is that in which the natural frequency is determined by nonlinear
tearing mode theory: in other words, the natural frequency is specified by Eq.~(\ref{e2}).

The upper panel of Fig.~\ref{fig5} shows the natural frequencies of the various  $n=3$ tearing modes 
 that are resonant in the pedestal. It can be seen that the natural frequency of the $m=11$ harmonic has its main zero crossing in the middle
of the first ELM-suppression window. Likewise, the natural frequency of the $m=10$ harmonic has its main zero crossing in the middle of the
second ELM-suppression window. However, the natural frequency of the $m=9$ harmonic has its main zero crossing slightly
before the third ELM-suppression window. The lower panel of Fig.~\ref{fig5} shows the actual phase velocities of the various $n=3$
tearing modes that are resonant in the pedestal. It can be seen that if the magnitude of the natural frequency falls below about
$5\,{\rm krad/s}$ then the associated tearing mode locks to the RMP (i.e., its phase velocity becomes zero).

\subsection{Case 2: $\varpi=\varpi_{EB}$}

\subsection{Case 3: $\varpi=\varpi_{\rm linear}$}

@@

\bibitem{nl1} R.~Fitzpatrick, and F.L.~Waelbroeck, Phys.\ Plasmas {\bf 12}, 022307 (2005).

\bibitem{nl2} R.~Fitzpatrick, Phys.\ Plasmas {\bf 25}, 042503 (2018).

\bibitem{nl3} R.~Fitzpatrick, Phys.\ Plasmas {\bf 25}, 112505 (2018).

@@

\bibitem{icoil} G.L.~Jackson, P.M.~Anderson, J.~Bialek, W.P.~Cary, G.L.~Campbell, A.M.~Garofalo,  R.~Hatcher, A.G.~Kellman, R.J.~La\,Haye, A.~Nagy, {\em et al.},  
Proc.\ 30th EPS Conf.\ on Controlled Fusion and Plasma Physics. (St.\ Petersburg, Russia, 2003) CD-ROM, P-4.47.

@@

\bibitem{neutral} P.~Monier-Garbet, K.H.~Burrell, F.L.~Hinton, J.~Kim, X.~Garbet, and R.J.~Groebner,   
Nucl.\ Fusion  {\bf 37}, 403 (1997).

\bibitem{llama} A.~Rosenthal, and F.M.~Laggner, private communication (2020).

\bibitem{solps} S~Wiesen, D.~Reiter, V.~Kotov, M.~Baelmans, W.~Dekeyser, A.S.~Kukushkin, S.W.~Lisgo, R.A.~Pitts, V.~Rozhansky, G.~Saibene I.~Veselova, and S.~Voskoboynikov,
J.\ Nucl.\ Mat.\ {\bf 468}, 480 (2015).

\bibitem{gim} R.~Fitzpatrick, C.G.~Gimblett, and R.J.~Hastie, 
Plasma Phys.\ Control.\ Fusion {\bf 34}, 161 (1992).

@@

\appendix
\section{Linear Versus Nonlinear Tearing Mode Theory in Tokamak Plasmas}\label{appa}
\subsection{Introduction}
The discussion in this Appendix outlines certain facts pertaining
to tearing modes in tokamak plasmas that ought to be common knowledge in the 
magnetic fusion community (but appear not to be).

\subsection{General Considerations}\label{general}
Consider a conventional tokamak plasma equilibrium. To lowest order, the equilibrium is governed by force balance,
\begin{equation}\label{ea}
-\nabla p + {\bf j}\times {\bf B} \simeq {\bf 0},
\end{equation}
and
flux freezing,
\begin{equation}\label{eb}
{\bf E} + {\bf V}\times {\bf B} \simeq {\bf 0}.
\end{equation}
This is the case because the other terms in the plasma equation of motion,
(\ref{ea}), and the plasma Ohm's law, (\ref{eb}), are much
smaller in magnitude than the retained terms. (Here, $p$ is the
scalar pressure, ${\bf j}$ the current density, ${\bf B}$ the
magnetic field strength, ${\bf E}$ the electric field strength, and ${\bf V}$ the plasma velocity). If we average the full plasma equation of motion and the full Ohm's law, as well as the full continuity and energy evolution equations, over magnetic flux-surfaces
then the dominant terms in these equations are annihilated, and
we end up with a set of $1\tfrac{1}{2}$-D evolution equations,
according to which the magnetic field and plasma current
evolve on the relatively long resistive timescale, $\tau_R$,
whereas the density and temperature evolve on a somewhat
shorter transport timescale.\cite{transp,gim}

A linear tearing mode is a helical instability of a tokamak plasma equilibrium that is related to a shear-Alfv\'{e}n wave.\cite{haz} The mode resonates with the equilibrium magnetic field at
the so-called rational magnetic flux-surface at which the
shear-Alfv\'{e}n velocity is zero. If $q$ is the safety-factor, $m$ the poloidal mode number, and $n$ the
toroidal mode number, then the rational
flux-surface is characterized by $q=m/n$. A thin resistive layer forms
around the rational surface that permits the reconnection of magnetic flux on a much faster timescale than the global
resistive evolution timescale, $\tau_R$, resulting in the formation of a helical magnetic island chain. Let the Alfv\'{e}n time,
$\tau_A$, be the
typical timescale on which a (compressional) Alfv\'{e}n
wave traverses the plasma.
In conventional tokamak plasmas, $\tau_A\ll \tau_R$. In fact,
the magnetic Lundquist number, $S\equiv \tau_R/\tau_A$,  typically exceeds $10^{\,8}$ in present-day tokamak plasmas,
and will likely exceed $10^{\,10}$ in ITER plasmas.
 It turns out that the width of the resistive layer is typically a factor $S^{\,-2/5}$ smaller than the
plasma minor radius, whereas magnetic flux is reconnected on
the hybrid timescale $\tau_A\ll \tau_A^{2/5}\,\tau_R^{3/5}\ll \tau_R$.\cite{fkr} Note that  the
structure of the tearing mode  outside the resistive layer is simply governed by a combination of flux-freezing and perturbed force balance.

@@

Figure~\ref{fig14} also sketches the
typical plasma current and vorticity patterns in the nonlinear regime. Observe that there is now a considerable distinction between an X-point and an O-point. The reason for this is that the island is sufficiently wide that the plasma cannot diffuse across it resistively rapidly enough to
avoid being trapped inside the magnetic separatrix. Indeed, both the regions inside and outside the magnetic separatrix
are governed by a combination of flux-freezing and perturbed force balance, whereas the resistive
layer, within which the plasma can easily slip through the magnetic field,  is diverted onto the magnetic separatrix. Moreover, as
the island grows, the width of the separatrix layer becomes
increasing negligible with respect to the island width. It
turns out that the separatrix layer does not significantly affect the evolution of a nonlinear magnetic island chain. Indeed,
the celebrated Rutherford island width evolution equation,
which governs the evolution of the island width, is simply a
$1\tfrac{1}{2}$-D evolution equation for a helical magnetic
equilibrium localized around the rational surface.\cite{ruth}
The Alfv\'{e}n time plays no role in this evolution (any more
than it plays a role in the $1\tfrac{1}{2}$-D evolution of
the global magnetic field). Likewise, the equation that
determines the phase velocity of a nonlinear magnetic island chain (see Ref.~\onlinecite{nl2}, and references therein) is essentially an
expression of the $1\tfrac{1}{2}$-D evolution of the plasma
flow profiles across the island region. As before, the  Alfv\'{e}n time plays no role in this evolution (any more
than it plays a role in the $1\tfrac{1}{2}$-D evolution of
the global plasma flow profiles).

It follows, from the previous discussion, that the nonlinear evolution of a tearing mode in a tokamak plasma has far more in common with the $1\tfrac{1}{2}$-D
evolution of the global plasma equilibrium than it does with
conventional linear tearing mode physics. In particular, the Alfv\'{e}n time is
an irrelevant timescale in nonlinear tearing mode theory, and
is also  very much shorter than the timescales on which
physical quantities of interest actually evolve. It should be noted that linear tearing layers in present-day tokamak plasmas are sufficiently thin that by the time a tearing mode is detectable it is already in the nonlinear regime (i.e., the island width exceeds the linear layer width).\cite{wesson} It, therefore, makes very little sense to
attempt to model nonlinear tearing mode evolution in tokamak plasmas using a toroidal, nonlinear, MHD code. The reason for this is obvious: employing  an MHD code necessarily introduces the very short Alfv\'{e}n time into the calculation,
but this timescale is actually irrelevant to the problem in hand. In fact, this is exactly the same reason that MHD codes are not usually used to reconstruct global plasma equilibria, or to model their time evolution.
